# BIOL 4800/6220 Fall 2024 R Tutorial

## Practical COmputing Fall 2024

```{r}
setwd("C:/Users/msbre/praccomp2024/")
```

### R Basics

#### Introduction

* R as a calculator

```{r}
1 + 100
```

```{r}
2^3
```

```{r}
7/2
```

```{r}
4-1
```

```{r}
2*3
```

```{r}
3+5*2
```

```{r}
(3+5)*2
```

```{r}
2e3
```

```{r}
2/10000
```

```{r}
log(1)
```

```{r}
exp(1)
```

* Boolean Conditionals

```{r}
1 == 1
```

```{r}
2 == 1
```

```{r}
2 > 1
```

```{r}
2 <= 1
```

```{r}
1 != 2
```

* Variables

```{r}
x <- 10
x
```

```{r}
x + x
```

```{r}
x - 5
```

```{r}
y <- x^x
```

```{r}
y
```

* Vectorization

```{r}
1:5
```

```{r}
2^(1:5)
```

```{r}
v <- 1:5
```

```{r}
2^v
```

```{r}
log(v)
```

* Environment

```{r}
ls()
```

```{r}
rm(v)
```

```{r}
ls()
```

#### Package Management

```{r}
#installed.packages()
```

```{r}
#install.packages("vegan", dependencies = TRUE)
library(vegan)
```

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("DESeq2")
```

```{r}
#library(DESeq2)
```

#### Getting Help

**Search the Web**

```{r}
?plot
```

```{r}
help(plot)
```

```{r}
vignette("FAQ-vegan")
```

```{r}
citation("vegan")
```

#### Data and their formats

```{r}
cats <- data.frame(coat = c("calico", "black", "tabby"),
                   weight = c(2.1, 5, 3.2),
                   likes.string = c(TRUE, FALSE, TRUE))
```

```{r}
cats
```

```{r}
getwd()
```

```{r}
write.csv(cats, "cats.csv")
```

```{r}
cats <- read.csv("cats.csv", header=TRUE, row.names = 1)
```

```{r}
cats
```

```{r}
cats$weight
```

```{r}
cats[, 2]
```

```{r}
cats[,"weight"]
```

```{r}
cats[c(1,3),c(2:3,1)]
```

```{r}
(cats.log.weight <- log(cats$weight))
```

```{r}
cbind(cats,cats.log.weight)
```

```{r}
(cats2 <- cbind(cats,cats.log.weight))
```

```{r}
cats
```

```{r}
paste("My cat is ", cats$coat, ", and it weighs ", cats$weight, " kg.", sep = "")
```

##### Data Types

```{r}
typeof(cats)
```

```{r}
class(cats)
```

```{r}
typeof(cats$coat)
typeof(cats$weight)
typeof(cats$likes.string)
```

```{r}
typeof(1i)
```

```{r}
likes.pop <- c(1,0,1,1,0)

typeof(likes.pop)
```

```{r}
as.logical(likes.pop)
```

```{r}
names <- c("Iggy","Bitey","Garfield")

cats$names <- names

cats
```

```{r}
(names2 <- c(names, "Spike"))
```

```{r}
(names3 <- c("Fluffy", names2))
```

```{r}
1:10
```

```{r}
(z <- seq(20, 150, by=0.5))
```

```{r}
length(z)
```

```{r}
length(cats)
```

```{r}
length(cats$coat)
```

```{r}
dim(cats)
```

```{r}
typeof(z)
```

```{r}
class(z)
```

```{r}
head(z, n=20)
```

```{r}
tail(z, n=40)
```

```{r}
sort(z, decreasing = TRUE)
```

```{r}
unique(z)
```

##### Data Frames

```{r}
cats
```

```{r}
cats$weight

typeof(cats$weight)
```

```{r}
str(cats$weight)
```

```{r}
coats <- c("tabby", "tortoiseshell", "tortoiseshell", "black", "tabby")
```

```{r}
(categories <- factor(coats))
```

```{r}
class(categories)
```

```{r}
str(categories)
```

##### Lists

```{r}
(list.example <- list(title="Numbers", numbers = seq(1,20,by=5), data=TRUE))
```

```{r}
list.example$title
```

```{r}
data.frame(list.example)
```

##### Matricies

```{r}
(matrix.example <- matrix(0, nrow = 3, ncol = 5))
```

```{r}
class(matrix.example)
```

```{r}
typeof(matrix.example)
```

```{r}
dim(matrix.example)
```

```{r}
nrow(matrix.example)
ncol(matrix.example)
```

```{r}
data.frame(matrix.example)
```

##### Subsetting

```{r}
(p <- c(2.3,6.9,4.0,23,1))
```

```{r}
names(p) <- c("a","b","c","d","e")
p
```

```{r}
typeof(p)
class(p)
```

```{r}
p[1]
```

```{r}
p[2:4]
```

```{r}
p[c(1,5)]
```

```{r}
p[c(1,1,1,3,5,5)]
```

```{r}
p[6]
```

```{r}
(q <- p[-3])
```

```{r}
p[-(2:4)]
```

```{r}
p[c("a", "e")]
```

```{r}
p[c(T,F,T,F,T)]
```

##### Factors

```{r}
(f <- factor(c("a","b","c","d","e")))
```

```{r}
f[f == "a"]
```

```{r}
f[1:3]
```

```{r}
(f2 <- factor(c("a","a","b","c","c")))
```

```{r}
f2[f2 == "a"]
```

```{r}
f2[f2 %in% c("a","c")]
```

##### Matricies Revisited

```{r}
set.seed(4398)

m <- matrix(rnorm(6,4,0.3), nrow=6, ncol=4)
m
```

```{r}
hist(rnorm(10000,4,0.3))
```

```{r}
m[3:4, c(3,1)]
```

##### Lists Revisited

```{r}
xlist <- list(a="BIOL48006220", b= rnorm(16,75,10), data="Grade")
xlist
```

##### Data Frames Revisited

```{r}
(gp <- read.csv("data/forest_area_sq_km.csv"))
```

```{r}
head(gp, n=10)
```

```{r}
head(gp[4], n=3L)
```

```{r}
nrow(gp)
```

```{r}
head(gp[["country"]], 10L)
```

```{r}
gp[c(1,3), c(1,5:10)]
```

```{r}
gp[ which(gp$X1989 <= 100 & gp$X2019 >= 100), ]
```

#### Conditionals and Flow

```{r}
n <- 1

if (n < 10) {
  print("n is less than 10.")
} else if (n > 10) {
  print("n is greater than 10.")
} else {
  print("n is equal to 10.")
}
```

```{r}
for (n in seq(0,20)) {
  if (n < 10) {
    print("n is less than 10.")
  } else if (n > 10) {
    print("n is greater than 10.")
  } else {
    print("n is equal to 10.")
  }
}
```

```{r}
for (n in rnorm(20,10,3)) {
  if (n < 10) {
    print(paste0(n," is less than 10."))
  } else if (n > 10) {
    print(paste0(n," is greater than 10."))
  } else {
    print(paste0(n," is equal to 10."))
  }
}
```

```{r}
g <- 0

while (g <= 10) {
  print(paste0(g," is less than or equal to 10."))
  g <- g+1
}
```

#### Plotting and Figures

```{r}
gap <- read.csv("~/praccomp2024/data/gapminder_all.csv")
gap
```

```{r}
plot(gap$gdpPercap_1982 ~ gap$pop_1982)
```

```{r}
#install.packages("ggplot2", dependencies=TRUE)
library(ggplot2)
```

```{r}
ggplot(data=gap, mapping=aes(y=gdpPercap_1982, x=pop_1982)) +
  geom_point()
```

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/main/episodes/data/gapminder_data.csv")

gapminder
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) +
  geom_point(aes(color=continent))
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) +
  geom_point(aes(color=continent), alpha=0.25) + 
  scale_x_log10() +
  geom_smooth(method = lm, color="black", fill="grey") +
  theme_classic()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year,y=lifeExp, by=country)) + 
  geom_line(mapping=aes(color=continent)) +
  geom_point() +
  theme_classic()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) +
  geom_point(color="darkblue", alpha=0.25, size=1) + 
  scale_x_log10() +
  geom_smooth(method = lm, color="deeppink", fill="chartreuse", size=0.5) +
  theme_classic()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) +
  geom_point(color="darkblue", alpha=0.25, size=1) + 
  geom_smooth(method = loess, color="deeppink", fill="chartreuse", size=0.5) +
  theme_classic()
```

```{r}
unique(gapminder$continent)
```

```{r}
asia <- gapminder[gapminder$continent == "Asia", ]
```

```{r}
ggplot(data=asia, mapping = aes(x=year, y=lifeExp)) +
  labs(
    x = "Year",
    y = "Life Expectency (years)",
    title = "Life expectency over time in Asian countries"
  ) +
  geom_line(color="red") +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle=90))
```

```{r}
ggsave("data/AsiaLifeExp.png")
```

```{r}
write.csv(asia, file ="data/asia.csv")
write.table(gapminder, file="data/gapminder_web_20240930.csv", sep=",")
```

###### Fancy Plots

```{r}
#install.packages(c("ggridges", "viridis", "hrbrthemes"), dependencies=TRUE)
```

```{r}
library(ggridges)
library(viridis)
library(hrbrthemes)
```

```{r}
ggplot(lincoln_weather, aes(x = `Mean Temperature [F]`, y = `Month`, fill = stat(x))) +
  labs(title = "Temperature in Lincoln, NE in 2016") + 
  geom_density_ridges_gradient(scales = 3, rel_min_height=0.01) +
  scale_fill_viridis_c(name = "Temp. [F]", option = "C") + 
  theme_classic()
```

```{r}
#install.packages(c("ggstatsplot", "tidyverse", "palmerpenguins"))
```

```{r}
library(tidyverse)
```

```{r}
data("penguins", package = "palmerpenguins")
penguins
```

```{r}
(penguins2 <- drop_na(penguins))
```

```{r}
ggplot(penguins2, aes(x=`bill_length_mm`, y=`bill_depth_mm`)) +
  geom_point() +
  geom_smooth(method="lm")
```

```{r}
ggplot(penguins2, aes(x=`bill_length_mm`, y=`bill_depth_mm`, color=`species`)) +
  geom_point() +
  geom_smooth(method="lm")
```

```{r}
#install.packages("rlang", dependencies=T)
library(ggstatsplot)

plt <- ggbetweenstats(
  data = penguins2,
  x=species,
  y=bill_length_mm
)

plt
```

```{r}
spider_data <- read.csv("data/spider_data_20221101.csv")
spider_data
```

```{r}
library(tidyverse)

spider_data %>%
  filter(!is.na(family)) %>%
  mutate(family = fct_rev(fct_infreq(family))) %>%
  ggplot(aes(x = family)) +
  geom_bar() +
  coord_flip() +
  #theme_classic() +
  theme(
    legend.position="none",
    panel.spacing = unit(0.1, "lines"),
    axis.text.y = element_text(size = 3)
  )
```

```{r}
library(ggridges)
library(ggplot2)
library(viridis)

spider_data %>%
  filter(!is.na(family)) %>%
  dplyr::count(family, sort=T) %>%
  slice(1:20) %>%
  left_join(spider_data) %>%
  
  ggplot(aes(x = `year`, y = reorder(family, n), color = `family`, fill = `family`)) +
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 25) +
  labs(title = "Spider species descriptions over time by family") +
  xlab(label = "Species Description Year") + 
  ylab(label = "Spider Family") +
  theme_ridges() +
  theme(
    legend.position="none",
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size = 8)
  )
```

```{r}
head(penguins2)
```

```{r}
penguins3 <- with(penguins2, cbind(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g))
```

```{r}
penguin_pca <- princomp(penguins3)
summary(penguin_pca)
```

```{r}
loadings(penguin_pca)
```

```{r}
biplot(penguin_pca)
```

```{r}
#install.packages("devtools")
library(devtools)

#install_github("vqv/ggbiplot")
library(ggbiplot)

ggbiplot(penguin_pca,
         groups = penguins2$species, ellipse = TRUE, scale. = TRUE) +
  scale_color_discrete(name = "")
```

```{r}
# Get data:
#install.packages("gapminder", dependencies = T)
#install.packages("gganimate", dependencies = T)
#install.packages("gifski", dependencies = T)
#install.packages("av", dependencies = T)

library(gapminder)
 
# Charge libraries:
library(ggplot2)
library(gganimate)
library(gifski)
 
# Make a ggplot, but add frame=year: one image per year
g <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_x_log10() +
  theme_bw() +
  # gganimate specific bits:
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')

print(g)

# Save at gif:
anim_save(filename = "data/271-ggplot2-animated-gif-chart-with-gganimate1.gif", animation = g)
```

### Statistics

```{r}
rand <- rnorm(50, 30, 8)
rand
```

```{r}
sum(rand)
```

```{r}
sum(rand)/length(rand)
```

```{r}
mean(rand)
```

```{r}
sort(rand)
```

```{r}
min(rand)
max(rand)
```

```{r}
cumsum(rand)
```

```{r}
diff(rand)
```

```{r}
sqrt(rand)
```

```{r}
log(rand)
```

```{r}
#things <- scan() # enter 1 2 3 4 5

things <- c(1,2,3,4,5) # use scan() line above if you are entering values manually
```

```{r}
things
```

#### Analyses

##### Univariate Statistics

1. Categorical Data
  - Barplots
  
```{r}
beer <- c(3,4,1,1,3,4,3,3,1,3,2,1,2,1,2,3,2,3,1,1,1,1,4,3,1)
length(beer)

table(beer)
```

```{r}
barplot(table(beer), xlab="Beer", ylab="Frequency", col="blue")
```

```{r}
barplot(table(beer)/length(beer), xlab="Beer", ylab="Frequency", col="blue")
```

```{r}
pie(table(beer), main = "Beer preference by students (n=25)")
```

2. Numerical Data

```{r}
rand
stem(rand)
```

```{r}
stripchart(rand)
```

3. Measures of center

```{r}
mean(rand)
```

```{r}
median(rand)
```

```{r}
which(table(rand) == max(table(rand)))
```

4. Variation

  - Range
  
```{r}
range(rand)
```

  - Variance
  
```{r}
var(rand)
```
```{r}
sd(rand)
```

  - IQR

```{r}
IQR(rand)
```

  - z-scores

```{r}
scale(rand)
```

  - Summary
  
```{r}
summary(rand)
```

5. Plots

  - Histograms

```{r}
hist(rand, breaks = 15, prob = TRUE)
```

  - Box plots
  
```{r}
boxplot(rand)
```

```{r}
#install.packages("vioplot")
library(vioplot)

vioplot(rand, col="blue")
```

##### Bivariate and multivariate statistics

  - Plotting and regression
    - Box plotting
    
```{r}
boxplot(gapminder$lifeExp~gapminder$continent, col = "darkorchid4")
```

    
```{r}
vioplot(gapminder$lifeExp~gapminder$continent, col="salmon")
```

```{r}
summary(rand)

t.test(rand, mu = 30)
```

```{r}
(spid.gen <- read.csv("data/spider_genitalia.csv", header=TRUE))
```

```{r}
vioplot(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat)
```

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb)
bulb.reg <- lm(spid.gen$left.bulb ~ spid.gen$right.bulb)
abline(bulb.reg)
```

```{r}
summary(bulb.reg)
```

```{r}
sqrt(0.5115)
```

```{r}
cor(spid.gen$left.bulb, spid.gen$right.bulb, method = "spearman")
```

```{r}
t.test(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat)
```

```{r}
wilcox.test(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat)
```

```{r}
ks.test(rand, "pnorm")
```

```{r}
shapiro.test(rand)
```

```{r}
qqnorm(rand, pch=16, col="red")
qqline(rand, lty=2, col="darkblue")
```


```{r}
summary(lm(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat))
```

```{r}
summary(aov(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat))
```

```{r}
habitat <- c("A", "A", "A", "A", "A", "B", "B", "B", "B", "B", "C", "C", "C", "C", "C")

parload <- c(10, 12, 8, 11, 14, 14, 16, 10, 9, 12, 18, 19, 20, 21, 22)

parhab <- cbind(habitat, parload)

parhab <- data.frame(parhab)

parhab$habitat <- as.factor(parhab$habitat)
parhab$parload <- as.numeric(parhab$parload)

parhab
```

```{r}
parhab.aov <- aov(parhab$parload ~ parhab$habitat)

summary(parhab.aov)
```

```{r}
(parhab.tukey <- TukeyHSD(parhab.aov))
```

```{r}
library(ggstatsplot)
ggbetweenstats(parhab, x = habitat, y = parload)

```

```{r}
(pvals <- c(0.05, 0.005, 0.0005, 0.00005, 0.000005))

p.adjust(pvals, method = "bonferroni")

p.adjust(pvals, method = "fdr")
```

```{r}
gapminder.2007 <- gapminder[which(gapminder$year == 2007), ]

gapminder.2007.pop.gdpPercap.lifeExp.continent <- lm(gapminder.2007$pop ~ gapminder.2007$gdpPercap + gapminder.2007$lifeExp + gapminder.2007$continent)

summary(gapminder.2007.pop.gdpPercap.lifeExp.continent)
```


```{r}
#install.packages("MuMIn", dependencies = TRUE)
library(MuMIn)

# Example from Burnham and Anderson (2002), page 100:

#  prevent fitting sub-models to different datasets
oop <- options(na.action = "na.fail")

dd <- dredge(gapminder.2007.pop.gdpPercap.lifeExp.continent)
subset(dd, delta < 2)

dd
```

```{r}
# Visualize the model selection table:
 if(require(graphics)) { 
par(mar = c(3,5,6,4))
plot(dd, labAsExpr = TRUE)
 } 
```

```{r}
# Model average models with delta AICc < 2
model.avg(dd, subset = delta < 2)

#or as a 95% confidence set:
model.avg(dd, subset = cumsum(weight) <= .95) # get averaged coefficients

#'Best' model
summary(get.models(dd, 1)[[1]])
```

### Geospatial Mapping

```{r}
#install.packages("maps", dependencies = T)

library(maps)

#install.packages(c("cowplot", "googleway", "ggplot2", "ggrepel", "ggspatial", "libwgeom", "sf", "rnaturalearth", "rnaturalearthdata", "rgbif", "terra", "geodata", "sdmpredictors", "stringr", "stars", "elevatr", "tigris", "data.table"))

#devtools::install_github("yutannihilation/ggsflabel")

library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(rgbif)
library(terra)
library(geodata)
library(sdmpredictors)
library(stringr)
library(raster)
library(ggspatial)
library(stars)
library(elevatr)
library(tigris)
library(ggsflabel)
library(data.table)
```

```{r}
geom_relief <- function(mapping = NULL, data = NULL,
                        stat = "identity", position = "identity",
                        ...,
                        raster = TRUE,
                        interpolate = TRUE,
                        na.rm = FALSE,
                        show.legend = NA,
                        inherit.aes = TRUE) {
   ggplot2::layer(
      data = data,
      mapping = mapping,
      stat = stat,
      geom = GeomRelief,
      position = position,
      show.legend = show.legend,
      inherit.aes = inherit.aes,
      params = list(
         raster = raster,
         interpolate = interpolate,
         na.rm = na.rm,
         ...
      )
   )
}

GeomRelief <- ggplot2::ggproto("GeomRelief", GeomTile,
  required_aes = c("x", "y", "z"),
  default_aes = ggplot2::aes(color = NA, fill = "grey35", size = 0.5, linetype = 1,
                             alpha = NA, light = "white", dark = "gray20", sun.angle = 60),
  draw_panel = function(data, panel_scales, coord, raster, interpolate) {
     if (!coord$is_linear()) {
        stop("non lineal coordinates are not implemented in GeomRelief", call. = FALSE)
     } else {
        coords <- as.data.table(coord$transform(data, panel_scales))
        
        # This is the only part that's actually new. The rest is essentially 
        # copy-pasted from geom_raster and geom_tile
        coords[, sun.angle := (sun.angle + 90)*pi/180]
        coords[, dx := .derv(z, x), by = y]
        coords[, dy := .derv(z, y), by = x]
        coords[, shade := (cos(atan2(-dy, -dx) - sun.angle) + 1)/2]
        coords[is.na(shade), shade := 0]
        coords[, fill := .rgb2hex(colorRamp(c(dark, light), space = "Lab")(shade)),
               by = .(dark, light)]
        
        # From geom_raster and geom_tile
        if (raster == TRUE){
           if (!inherits(coord, "CoordCartesian")) {
              stop("geom_raster only works with Cartesian coordinates", call. = FALSE)
           }
           # Convert vector of data to raster
           x_pos <- as.integer((coords$x - min(coords$x)) / resolution(coords$x, FALSE))
           y_pos <- as.integer((coords$y - min(coords$y)) / resolution(coords$y, FALSE))
           
           nrow <- max(y_pos) + 1
           ncol <- max(x_pos) + 1
           
           raster <- matrix(NA_character_, nrow = nrow, ncol = ncol)
           raster[cbind(nrow - y_pos, x_pos + 1)] <- alpha(coords$fill, coords$alpha)
           
           # Figure out dimensions of raster on plot
           x_rng <- c(min(coords$xmin, na.rm = TRUE), max(coords$xmax, na.rm = TRUE))
           y_rng <- c(min(coords$ymin, na.rm = TRUE), max(coords$ymax, na.rm = TRUE))
           
           grid::rasterGrob(raster,
                            x = mean(x_rng), y = mean(y_rng),
                            width = diff(x_rng), height = diff(y_rng),
                            default.units = "native", interpolate = interpolate
           )
           
        } else {
           ggplot2:::ggname("geom_rect", grid::rectGrob(
              coords$xmin, coords$ymax,
              width = coords$xmax - coords$xmin,
              height = coords$ymax - coords$ymin,
              default.units = "native",
              just = c("left", "top"),
              gp = grid::gpar(
                 col = coords$fill,
                 fill = alpha(coords$fill, coords$alpha),
                 lwd = coords$size * .pt,
                 lty = coords$linetype,
                 lineend = "butt"
              )
           ))
           
        }
     }
  }
)

rect_to_poly <- function(xmin, xmax, ymin, ymax) {
   data.frame(
      y = c(ymax, ymax, ymin, ymin, ymax),
      x = c(xmin, xmax, xmax, xmin, xmin)
   )
}

.rgb2hex <- function(array) {
   rgb(array[, 1], array[, 2], array[, 3], maxColorValue = 255)
}


.derv <- function(x, y, order = 1, cyclical = FALSE, fill = FALSE) {
   N <- length(x)
   d <- y[2] - y[1]
   if (order >= 3) {
      dxdy <- .derv(.derv(x, y, order = 2, cyclical = cyclical, fill = fill),
                    y, order = order - 2, cyclical = cyclical, fill = fill)
   } else {
      if (order == 1) {
         dxdy <- (x[c(2:N, 1)] - x[c(N, 1:(N-1))])/(2*d)
      } else if (order == 2) {
         dxdy <- (x[c(2:N, 1)] + x[c(N, 1:(N-1))] - 2*x)/d^2
      }
      if (!cyclical) {
         if (!fill) {
            dxdy[c(1, N)] <- NA
         }
         if (fill) {
            dxdy[1] <- (-11/6*x[1] + 3*x[2] - 3/2*x[3] + 1/3*x[4])/d
            dxdy[N] <- (11/6*x[N] - 3*x[N-1] + 3/2*x[N-2] - 1/3*x[N-3])/d
         }
      }
      
   }
   return(dxdy)
}
```


```{r}
theme_set(theme_bw())

world <- ne_countries(scale = "medium", returnclass = "sf")

world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

ggplot(data = world) +
  geom_sf() +
  #geom_sf(data = water, fill = "lightblue") +
  geom_text(data= world_points,
            aes(x=X, y=Y, label=name),
            color = "darkblue", 
            fontface = "bold", 
            size = 2,
            check_overlap = TRUE)
```

```{r}
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet))

world_relief <- get_elev_raster(locations = world, z = 1, clip = "locations")

#r <- raster(xmn = -105, xmx = -50, ymn = 0, ymx = 55, nrow = 100, ncol = 100)
#r[]=runif(100*100)

plot(world_relief)
plot(ne_coastline(), add=TRUE)
#plot(r, add=TRUE)
```

```{r}
nc_counties <- counties(
  state = 'North Carolina', cb = FALSE, progress_bar = TRUE)
```

```{r}
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet))

# Use elevatr to get elevation data
nc_elevations <- elevatr::get_elev_raster(
  locations = nc_counties, z = 8, clip = 'locations'
)

plot(nc_elevations)

# Convert elevation data to dataframe
nc_elevations <- as.data.frame(nc_elevations, xy = TRUE)
colnames(nc_elevations)[3] <- 'elevation'

# Remove rows with one or more NA's using complete.cases
nc_elevations <- nc_elevations[complete.cases(nc_elevations), ]
```

```{r}
water <- st_read("data/USA_Detailed_Water_Bodies.shp")

water <- st_transform(water, crs = st_crs(nc_counties))

#plot(water)

sf_use_s2(FALSE)

nc.water <- st_intersection(water, nc_counties)

sf_use_s2(TRUE)
```


```{r}
ggplot() +
  geom_relief(data = nc_elevations,
              aes(x = x, y = y, z = elevation, light = 'white', dark = 'grey20'), 
               raster = FALSE, interpolate = TRUE, sun.angle = 60) +
  geom_sf(data = nc.water, color = NA, fill = 'lightblue', size = 0.6, alpha = 1) +
  geom_sf(data = nc_counties, color = 'white', fill = NA) +
  geom_sf_text_repel(data = nc_counties, aes(label = NAME),
                     size = 3, bg.color = 'white', color = 'darkgreen') +
  coord_sf(datum = sf::st_crs(nc_counties), ) +
  labs(title = 'Land Elevations',
       subtitle = 'Counties of North Carolina',
       caption = 'Data source: https://registry.opendata.aws/terrain-tiles')
```

```{r}
Appalachioria.separanda_GBIF <- occ_data(scientificName = "Appalachioria separanda", hasCoordinate = TRUE, decimalLongitude = "-100, -60", decimalLatitude = "0, 50", limit = 10000)

Appalachioria.separanda_GBIF <- data.frame(Appalachioria.separanda_GBIF$data)

Appalachioria.separanda_GBIF
```

```{r}
states <- states()

zone <- counties(
  state = c('Virginia', 'West Virginia', 'Maryland', 'Pennsylvania', 'North Carolina'), cb = FALSE, progress_bar = TRUE)

#zone <- states[ which(states$STUSPS == "WV" | states$STUSPS == "MD" | states$STUSPS == "PA" | states$STUSPS == "VA" | states$STUSPS == "DC"),]

zone
```


```{r}
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet))

# Use elevatr to get elevation data
elevations <- elevatr::get_elev_raster(
  locations = zone, z = 8, clip = 'locations'
)

# Convert elevation data to dataframe
elevations <- as.data.frame(elevations, xy = TRUE)
colnames(elevations)[3] <- 'elevation'

# Remove rows with one or more NA's using complete.cases
elevations <- elevations[complete.cases(elevations), ]
```

```{r}
water <- st_read("data/USA_Detailed_Water_Bodies.shp")

water <- st_transform(water, crs = st_crs(zone))

#plot(water)

sf_use_s2(FALSE)

zone.water <- st_intersection(water, zone)

sf_use_s2(TRUE)
```

```{r}
ggplot() +
  geom_relief(data = elevations,
              aes(x = x, y = y, z = elevation, light = 'white', dark = 'grey20'), 
               raster = FALSE, interpolate = TRUE, sun.angle = 60) +
  geom_sf(data = zone.water, color = NA, fill = 'lightblue', size = 0.6, alpha = 1) +
  geom_sf(data = zone, color = NA, fill = NA) +
  #geom_sf_text_repel(data = zone, aes(label = NAME),
  #                   size = 3, bg.color = 'white', color = 'skyblue') +
  coord_sf(datum = sf::st_crs(zone)) +
  geom_point(data = Appalachioria.separanda_GBIF, 
             aes(x = decimalLongitude, y = decimalLatitude, color="red")) +
  labs(title = 'Land Elevations',
       subtitle = 'State in Zone of Interest',
       caption = 'Data source: https://registry.opendata.aws/terrain-tiles')

```


# ENM workbook

## Citation
- GBIF.org (12 October 2023) GBIF Occurrence Download  https://doi.org/10.15468/dl.3stbgp

## Download, read, and explore occurence data
```{r}
Appalachioria.separanda_GBIF <- occ_data(scientificName = "Appalachioria separanda", hasCoordinate = TRUE, decimalLongitude = "-100, -60", decimalLatitude = "0, 50", limit = 10000)

write.csv(Appalachioria.separanda_GBIF$data, file = "data/Appalachioria.separanda_GBIF.csv", row.names = FALSE)

Appalachioria.separanda_GBIF <- read.csv("data/Appalachioria.separanda_GBIF.csv")
```

```{r}
Appalachioria.separanda_GBIF
```

### Construct SDM

#### Obtain Environmental Layers

a. View available layers
```{r}
# Inspect the available datasets and layers
datasets <- list_datasets(terrestrial = TRUE, marine = FALSE)
#View(datasets)
layers <- list_layers(datasets)
#View(layers)
```

b. Choose and load layers
```{r}
# Load equal area rasters
layercodes <- c("WC_alt", "WC_bio1", "WC_bio2", "WC_bio3", "WC_bio4", "WC_bio5", "WC_bio6", "WC_bio7", "WC_bio8", "WC_bio9", "WC_bio10", "WC_bio11", "WC_bio12", "WC_bio13", "WC_bio14", "WC_bio15", "WC_bio16", "WC_bio17", "WC_bio18", "WC_bio19")
env <- load_layers(layercodes,  datadir = "sdm_data_layers")
#plot(env)

easternUS <- raster::crop(env, raster::extent(-100, -60, 0, 50))
plot(easternUS)
```

c. 
```{r}
# Compare correlations between predictors for Carribean
prettynames <- list(WC_alt="Altitude", 
                    WC_bio1="Annual mean temp", 
                    WC_bio2="Mean diurnal temperature range", 
                    WC_bio3="Isothermality", 
                    WC_bio4="Temperature seasonality", 
                    WC_bio5="Maximum temperature", 
                    WC_bio6="Minimum temperature", 
                    WC_bio7="Annual temperature range", 
                    WC_bio8="Mean temperature of wettest quarter", 
                    WC_bio9="Mean temperature of driest quarter", 
                    WC_bio10="Mean temperature of warmest quarter", 
                    WC_bio11="Mean temperature of coldest quarter", 
                    WC_bio12="Annual precipitation", 
                    WC_bio13="Precipitation of wettest month", 
                    WC_bio14="Precipitation of driest month", 
                    WC_bio15="Precipitation seasonality", 
                    WC_bio16="Precipitation of wettest quarter", 
                    WC_bio17="Precipitation of driest quarter", 
                    WC_bio18="Precipitation of warmest quarter", 
                    WC_bio19="Precipitation of coldest quarter")

easternUS_correlations <- pearson_correlation_matrix(easternUS)

p1 <- plot_correlation(easternUS_correlations, prettynames)

cowplot::plot_grid(p1, labels = "Eastern US", ncol = 1, nrow = 1)

print(correlation_groups(easternUS_correlations))
```

#### Choose occurence points for SDMs

```{r}
#summary(Appalachioria.separanda_GBIF)
```

```{r}
obs_data <- Appalachioria.separanda_GBIF[!is.na(Appalachioria.separanda_GBIF$decimalLatitude), ]
```

#### Crop environmental layers to fit occurences

```{r}
# Determine geographic extent of our data
max_lat <- ceiling(max(obs_data$decimalLatitude))
min_lat <- floor(min(obs_data$decimalLatitude))
max_lon <- ceiling(max(obs_data$decimalLongitude))
min_lon <- floor(min(obs_data$decimalLongitude))
# Store boundaries in a single extent object
geographic_extent <- raster::extent(x = c(min_lon, max_lon, min_lat, max_lat))

sample_extent <- geographic_extent * 1.25

sample_extent
```


```{r}
# Fetch occurrences and prepare for ???
points <- cbind(obs_data$decimalLongitude, obs_data$decimalLatitude)
#points <- spTransform(points)
occfile <- tempfile(fileext = ".csv")
write.csv(cbind(points, value=1), occfile)
#read.csv("occfile")
```

```{r}
# Download data with geodata's world function to use for our base map
world_map <- world(resolution = 3,
                   path = "data/")

# Crop the map to our area of interest
my_map <- crop(x = world_map, y = sample_extent)

# Plot the base map
plot(my_map,
     axes = TRUE, 
     col = "grey95")

# Add the points for individual observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       col = "olivedrab", 
       pch = 20, 
       cex = 0.75)
```

#### FIX!!!

```{r}
bioclim_data <- worldclim_global(var = "bio",
                                 res = 2.5,
                                 path = "data")
```

```{r}
# Make an extent that is 25% larger
sample_extent <- geographic_extent * 1.25

# Crop bioclim data to desired extent
bioclim_data <- crop(x = env, y = sample_extent)

# Plot the first of the bioclim variables to check on cropping
plot(bioclim_data[[1]])
```

```{r}
bioclim_data_SR <- as(bioclim_data, "SpatRaster")
```


```{r}
# Set the seed for the random-number generator to ensure results are similar
set.seed(20210707)

# Randomly sample points (same number as our observed points)
background <- spatSample(x = bioclim_data_SR,
                         size = 100000,    # generate 1,000 pseudo-absence points
                         values = FALSE, # don't need values
                         na.rm = TRUE,   # don't sample from ocean
                         xy = TRUE)      # just need coordinates

# Look at first few rows of background
head(background)
```
```{r}
# Plot the base map
plot(my_map,
     axes = TRUE, 
     col = "grey95")

# Add the background points
points(background,
       col = "grey30",
       pch = 2,
       cex = 0.1,)

# Add the points for individual observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       col = "firebrick", 
       pch = 20, 
       cex = 0.25)
```

```{r}
# Pull out coordinate columns, x (longitude) first, then y (latitude) from presences
presence <- obs_data[, c("decimalLongitude", "decimalLatitude")]
# Add column indicating presence
presence$pa <- 1

# Convert background data to a data frame
absence <- as.data.frame(background)
# Update column names so they match presence points
colnames(absence) <- c("decimalLongitude", "decimalLatitude")
# Add column indicating absence
absence$pa <- 0

# Join data into single data frame
all_points <- rbind(presence, absence)

#Convert column names to match bioclim
colnames(all_points) <- c("longitude", "latitude","pa")

# Reality check on data
head(all_points)
```

```{r}
bioclim_extract <- extract(x = bioclim_data_SR,
                           y = all_points[, c("longitude", 
                                              "latitude")],
                           ID = FALSE)
```

```{r}
# Add the point and climate datasets together
points_climate <- cbind(all_points, bioclim_extract)

# Identify columns that are latitude & longitude
drop_cols <- which(colnames(points_climate) %in% c("longitude", "latitude"))
drop_cols # print the values as a reality check
```

```{r}
# Remove the geographic coordinates from the data frame
points_climate <- points_climate[, -drop_cols]
```

```{r}
# Create vector indicating fold
#install.packages("predicts", dependencies = T)
library(predicts)

fold <- folds(x = points_climate,
              k = 5,
              by = points_climate$pa)

table(fold)
```

```{r}
testing <- points_climate[fold == 1, ]
training <- points_climate[fold != 1, ]
```

```{r}
# Build a model using training data
glm_model <- glm(pa ~ ., data = training, family = binomial())
```

```{r}
# Get predicted values from the model
glm_predict <- predict(bioclim_data_SR, glm_model, type = "response")

# Print predicted values
plot(glm_predict)
```


```{r}
# Use testing data for model evaluation
glm_eval <- pa_evaluate(p = testing[testing$pa == 1, ],
                        a = testing[testing$pa == 0, ],
                        model = glm_model,
                        type = "response")
```

```{r}
# Determine minimum threshold for "presence"
glm_threshold <- glm_eval@thresholds$max_spec_sens
```

```{r}
# Plot base map
plot(my_map, 
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(glm_predict > glm_threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = "olivedrab")

# And add those observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       col = "black",
       pch = "+", 
       cex = 0.75)

# Redraw those country borders
plot(my_map, add = TRUE, border = "grey5")
```

```{r}
glm_predict > glm_threshold
```

```{r}
# Plot base map
plot(my_map, 
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(glm_predict > glm_threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = c(NA, "olivedrab")) # <-- Update the values HERE

# And add those observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       #col = alpha("blue",0.5),
       col = "blue",
       pch = ".", 
       cex = 0.0000000000001
       )

# Redraw those country borders
plot(my_map, add = TRUE, border = "grey5")
```
```{r}
# Save PDF
pdf("data/Appalachioria.separanda_GBIF_GLM_Pred.pdf")

# Plot base map
plot(my_map, 
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(glm_predict > glm_threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = c(NA, "olivedrab")) # <-- Update the values HERE

# And add those observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       #col = alpha("blue",0.5),
       col = "blue",
       pch = ".", 
       cex = 0.0000000000001
       )

# Redraw those country borders
plot(my_map, add = TRUE, border = "grey5")

dev.off()
```

```{r}
# Layer citations
print(layer_citations(layercodes))
```


```{r}
glm_model
```

```{r}
geotiffs <- list.files(path = "data/climate/wc2.1_2.5m/", pattern = ".tif")

dir.create("data/ascii_files")

for (i in geotiffs) {
  f <- raster::raster(paste0("data/climate/wc2.1_2.5m/",i))
  r <- raster::readAll(f)
  #r <- raster(r)
  r <- raster::crop(r, sample_extent)
  writeRaster(r, paste0("data/ascii_files/",i,".asc"), format = "ascii", overwrite=TRUE)
}
```

```{r}
#install.packages("MIAmaxent", dependencies = T)
library(MIAmaxent)
library(dplyr)
library(tibble)

GBIF4MIAmaxent <- as.data.frame(cbind(Appalachioria.separanda_GBIF$decimalLongitude, Appalachioria.separanda_GBIF$decimalLatitude))

GBIF4MIAmaxent <- GBIF4MIAmaxent %>%
  add_column(Occurrence = "presence")

colnames(GBIF4MIAmaxent) <- c("POINT_X", "POINT_Y", "Occurrence")

GBIF4MIAmaxent <- GBIF4MIAmaxent %>%
  dplyr::select(Occurrence, everything())

GBIF4MIAmaxent

write.csv(GBIF4MIAmaxent, "data/GBIF4MIAmaxent.csv", row.names = FALSE)
```

```{r}
GBIFPO <- readData(
  occurrence = "data/GBIF4MIAmaxent.csv",
  contEV = "data/ascii_files/",
  maxbkg=20000,
  PA = FALSE
)

```

```{r}
EV1 <- raster::raster(list.files("data/ascii_files/", full.names=TRUE)[1])

PO <- read.csv("data/GBIF4MIAmaxent.csv")

plot(EV1, legend=FALSE)
points(PO$POINT_X, PO$POINT_Y, pch = 20, cex = 0.5, col = "blue")
```

```{r}
str(GBIFPO)
```

```{r}
for (n in list.files(path = "data/climate/wc2.1_2.5m/", pattern = ".tif")) {
  plotFOP(GBIFPO, stringr::str_replace_all(n, pattern = "_", replacement = "."), span = 0.75, intervals = 20)
}
```


```{r}
GBIFDVs <- deriveVars(GBIFPO, transformtype = c("L","M","D","HF","HR","T","B"))
```

```{r}
GBIFDVselect <- selectDVforEV(GBIFDVs$dvdata, alpha = 0.001, quiet = TRUE)
```

```{r}
GBIFEVselect <- selectEV(GBIFDVselect$dvdata, alpha = 0.001, 
                              interaction = TRUE)
```

```{r}
GBIFEVselect$selectedmodel$formula
```

```{r}
GBIFEVselect$selection[!duplicated(GBIFEVselect$selection$round), ]
```

```{r}
plot(GBIFEVselect$selection$round, GBIFEVselect$selection$Dsq, 
     xlab="round", ylab="Dsq")
```

```{r}
#GBIFmodeltext <- cbind.data.frame(GBIFEVselect$selection[!duplicated(GBIFEVselect$selection$round), ])[34,][,2]

#GBIFmodeltext

#GBIFmodel <- chooseModel(GBIFDVselect$dvdata, formula(paste0("~ ", GBIFmodeltext)))
```

```{r}
GBIFmodel <- GBIFEVselect$selectedmodel
```

```{r}
EVfiles <- c(list.files("data/ascii_files/", pattern = ".asc", full.names=TRUE))
EVstack <- raster::stack(EVfiles)
#names(EVstack) <- gsub("_", ".", gsub(".asc", "", basename(EVfiles)))
GBIFPreds <- projectModel(model = GBIFmodel,
                               transformations = GBIFDVs$transformations,
                               data = EVstack)
```

```{r}
plot(log2(GBIFPreds$output +1))

pdf("GBIF_MIAMAXENT_Pred_MaxEnt.pdf")
plot(log2(GBIFPreds$output +1))
dev.off()
```

```{r}
testAUC(model = GBIFmodel, transformations = GBIFDVs$transformations,
        data = GBIFPO)
```

```{r}
GBIFDVs.LR <- deriveVars(GBIFPO, algorithm = "LR")
```

```{r}
GBIFDVselect.LR <- selectDVforEV(GBIFDVs.LR$dvdata, alpha = 0.001,
                                 algorithm = "LR")

GBIFEVselect.LR <- selectEV(GBIFDVselect.LR$dvdata, alpha = 0.001,
                            algorithm = "LR")
```

```{r}
GBIFEVselect.LR$selection
```

```{r}
GBIFEVselect.LR$selectedmodel
```

```{r}
GBIFPreds.LR <- projectModel(model = GBIFEVselect.LR$selectedmodel,
                               transformations = GBIFDVs.LR$transformations,
                               data = EVstack)

pdf("GBIF_MIAMAXENT_Pred_LR.pdf")
plot(GBIFPreds.LR$output)
dev.off()

plot(GBIFPreds.LR$output)
```

```{r}
testAUC(GBIFEVselect.LR$selectedmodel, GBIFDVs.LR$transformations, GBIFPO)
```

```{r}
#install.packages("tidyterra")

library(tidyterra)

sf_use_s2(FALSE)

elevations4MaxEnt <- elevations %>% filter(x >= as.list(sample_extent)[1] & x <= as.list(sample_extent)[2] & y >= as.list(sample_extent)[3] & y <= as.list(sample_extent)[4])

ggplot() +
  geom_relief(data = elevations4MaxEnt,
              aes(x = x, y = y, z = elevation, light = 'white', dark = 'grey20'), 
               raster = FALSE, interpolate = TRUE, sun.angle = 60) +
  geom_sf(data = st_crop(zone.water, sample_extent), color = NA, fill = 'lightblue', size = 0.6, alpha = 1) +
  geom_sf(data = st_crop(zone, sample_extent), color = NA, fill = NA) +
  #geom_sf_text_repel(data = zone, aes(label = NAME),
  #                   size = 3, bg.color = 'white', color = 'skyblue') +
  geom_spatraster(data = GBIFPreds$output %>% filter(PRO >= 1), alpha = 0.5) +
  scale_fill_whitebox_c(palette = "muted", na.value = "transparent", guide = 'none') +
  coord_sf(datum = sf::st_crs(zone)) +
  geom_point(data = Appalachioria.separanda_GBIF, 
             aes(x = decimalLongitude, y = decimalLatitude, color="purple"),
             show.legend = FALSE) +
  labs(title = 'ENM of Appalachioria separanda',
       subtitle = 'Over DEM',
       caption = 'Data source: https://registry.opendata.aws/terrain-tiles') +
  theme(axis.text.x = element_text(angle=90))

sf_use_s2(TRUE)
```

```{r}
unloadNamespace("tidyterra")
```


### Tidyverse

```{r}
library(tidyverse)
library(palmerpenguins)

head(penguins)
```

```{r}
weather <- read.csv("https://pasta.lternet.edu/package/data/eml/knb-lter-pal/28/8/375b34051b162d84516ec2d02f864675")
```

- Structure

```{r}
str(penguins)
```

```{r}
str(weather)
```

- Reducing Data Set Dimensions

```{r}
colnames(penguins)
```

```{r}
dim(penguins)
```

```{r}
Torg_2007_penguins <- penguins %>% dplyr::select(species, island, year) %>%
  dplyr::filter(year == 2007 & island == "Torgersen")

Torg_2007_penguins
```

```{r}
is.na(penguins$sex)
```
```{r}
vioplot(penguins$body_mass_g)
```

```{r}
summary(penguins %>% filter(!is.na(sex) & body_mass_g >= 5000))
```

```{r}
penguins_bill.mass.ratio <- penguins %>% mutate(bill.mass.ratio = bill_length_mm/body_mass_g)

vioplot(penguins_bill.mass.ratio$bill.mass.ratio)
```

```{r}
penguins %>% filter(!is.na(body_mass_g)) %>%
  mutate(body_size_cat = case_when(
    body_mass_g >= 5000 ~ "big",
    body_mass_g <= 3500 ~ "small",
    body_mass_g <5000 & body_mass_g > 3500 ~ "medium",
    T ~ "ERROR"
))
```

```{r}
penguins %>% filter(!is.na(body_mass_g)) %>%
  mutate(body_size_cat = case_when(
    body_mass_g >= 5000 ~ "big",
    body_mass_g <= 3500 ~ "small",
    body_mass_g <5000 & body_mass_g > 3500 ~ "medium",
    T ~ "ERROR"
  )) %>%
  filter(!is.na(sex)) %>%
  dplyr::group_by(species, sex) %>%
  dplyr::summarize(mean_body_mass_g = mean(body_mass_g, na.rm=TRUE),
                   sd_body_mass_g = sd(body_mass_g, na.rm=TRUE))
```

- Manipulating Data

```{r}
str(weather)
```

```{r}
lubridate::yday("2024-10-28")
```

```{r}
weather %>% mutate(Date = ymd(Date)) %>% 
  mutate(Date = floor_date(Date, unit = 'month')) %>%
  group_by(Date) %>% 
  dplyr::summarise(mean(Temperature.Average..C.)) %>%
  
  ggplot(aes(x=Date, y=`mean(Temperature.Average..C.)`)) +
    geom_point(aes(color = year(Date))) + 
    scale_color_viridis_c() +
    geom_line() +
    geom_smooth(method="lm", color="red", linewidth = 0.5, fill = "orange")
```

- Joining data frames

```{r}
colnames(penguins)
```

```{r}
colnames(weather)
```

```{r}
(weather.sums <- weather %>%
  group_by(year(Date)) %>%
  dplyr::summarise(avg.temp = mean(Temperature.Average..C., na.rm=TRUE),
                   max.temp = max(Temperature.High..C., na.rm=TRUE),
                   min.temp = min(Temperature.Low..C., na.rm=TRUE)) %>%
  dplyr::rename(year = `year(Date)`)
)
```

```{r}
(penguins_weather.sums <- left_join(penguins, weather.sums, by='year'))
```

```{r}
penguins_weather.sums %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(mean(avg.temp))
```


```{r}
penguins_weather.sums %>%
  dplyr::filter(!is.na(sex)) %>%
  ggplot(aes(x=sex, y=body_mass_g, color = sex, fill = sex)) +
  geom_violin() +
  scale_color_manual(values = c("darkorange", "purple")) +
  scale_fill_manual(values = c("darkorange", "purple")) +
  facet_wrap( ~ species + year) +
  theme_dark() +
  labs(title = "Average penguin body size per species per year by sex",
       x = "Sex",
       y = "Body Mass (g)")
```


```{r}
penguins_weather.sums %>%
  dplyr::filter(!is.na(sex)) %>%
  dplyr::group_by(species, sex, year) %>%
  dplyr::summarize(mean.body.mass.g = mean(body_mass_g),
            mean.avg.temp.c = mean(avg.temp),
            mean.min.temp.c = mean(min.temp),
            mean.max.temp.c = mean(max.temp)) %>%
  ggplot(aes(x=mean.min.temp.c, y=mean.body.mass.g, color = sex)) +
  geom_point() +
  scale_color_manual(values = c("darkorange", "purple")) +
  geom_smooth(method="lm", se = FALSE) +
  facet_wrap( ~ species) +
  theme_classic() +
  labs(title = "Average penguin body mass by average annual minimum temperature",
       x = "Average minimum temperature (C)",
       y = "Average body mass (g)")
  
```



